---
title: "Bayesian Inference with NIMBLE"
output: slidy_presentation
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = FALSE)
```


# Simple linear regression in NIMBLE

```{r}
library(nimble)
```

# Simulate data

\begin{align*}
y_i \sim N(\beta_0 + \beta_1 x_i, \sigma^2)
\end{align*}

```{r}
set.seed(1)
n <- 100
x <- runif(n)
b0_true <- 0.3
b1_true <- 0.5
sigma_true <- 0.25
y <- rnorm(n, b0_true + b1_true*x, sigma_true)

plot(x, y)
```


# NIMBLE model syntax

```{r}
lmCode <- nimbleCode({
    
    # likelihood
    for(i in 1:n) {
        mu[i] <- b0 + b1*x[i] 
        y[i] ~ dnorm(mu[i], sd = sigma)
    }
    
    # priors
    b0 ~ dnorm(0, sd = 100)
    b1 ~ dnorm(0, sd = 100)
    sigma ~ dunif(0, 100)
    
})
```

# Sampling from the posterior (easy method)

After writing the model, we can obtain samples from the posterior. The most 
straightforward method for this is to use the nimbleMCMC function. This uses
the default MCMC algorithm.

```{r}
dataList <- list(y = y)
constantsList <- list(n = n, x = x)
initsList <- list(b0 = 1, b1 = 0, sigma = 1)

system.time({
    samples <- nimbleMCMC(lmCode,
                          data = dataList, 
                          constants = constantsList,
                          inits = initsList,
                          niter = 10000)
})


head(samples)

par(mfrow = c(1,3))
plot(samples[,'b0'], type = 'l')
abline(h = b0_true, col = 'red')
plot(samples[,'b1'], type = 'l')
abline(h = b1_true, col = 'red')
plot(samples[,'sigma'], type = 'l')
abline(h = sigma_true, col = 'red')
```


# Sampling from the posterior (customizable method)

We can also obtain samples using a longer method which provides the ability to 
customize the MCMC algorithm. Using this method, there are 5 steps:

1. Build the nimbleModel object

```{r}
### create nimble model object
lmModel <- nimbleModel(lmCode, 
                       data = dataList, 
                       constants = constantsList,
                       inits = initsList)
```


2. Create the MCMC configuration object

```{r}
lmConfigure <- configureMCMC(lmModel)
```

3. Build the MCMC object from the configuration

```{r}
lmMCMC <- buildMCMC(lmConfigure)
```

4. Compile the model and MCMC in C++

```{r}
lmCompiled <- compileNimble(lmModel, lmMCMC) 
```

5. Run the MCMC to obtain samples

```{r}
samples2 <- runMCMC(lmCompiled$lmMCMC, niter = 10000)

head(samples2)

par(mfrow = c(1,3))
plot(samples2[,'b0'], type = 'l')
abline(h = b0_true, col = 'red')
plot(samples2[,'b1'], type = 'l')
abline(h = b1_true, col = 'red')
plot(samples2[,'sigma'], type = 'l')
abline(h = sigma_true, col = 'red')
```

# Customizing the MCMC

Using the customizable method, we can customize the MCMC. For example, we could use a slice
sampler instead of a random walk sampler for sigma.

```{r}
lmConfigure$removeSamplers('sigma')
lmConfigure$addSampler('sigma',
                       type = 'slice')

lmConfigure

lmMCMC <- buildMCMC(lmConfigure)

lmCompiled <- compileNimble(lmModel, lmMCMC) 

samples3 <- runMCMC(lmCompiled$lmMCMC, niter = 10000)

head(samples3)

par(mfrow = c(1,3))
plot(samples3[,'b0'], type = 'l')
abline(h = b0_true, col = 'red')
plot(samples3[,'b1'], type = 'l')
abline(h = b1_true, col = 'red')
plot(samples3[,'sigma'], type = 'l')
abline(h = sigma_true, col = 'red')
```



# nimbleModels are objects


The nimbleModel object contains the current state of the model parameters.

```{r}
lmModel$b0
lmModel$b1
lmModel$sigma
```

nimbleModel objects also contain methods (functions). The calculate() and 
getLogProb() methods calculate the log posterior for the current state of the model.

```{r}
lmModel$getLogProb()

lmModel$calculate()

sum(dnorm(y, mean = lmModel$b0 + lmModel$b1*x, sd = lmModel$sigma, log = T)) +
    dnorm(lmModel$b0, 0, 100, log = T) + 
    dnorm(lmModel$b1, 0, 100, log = T) +
    dunif(lmModel$sigma, 0, 100, log = T)
```

When calculate() is called, the stored value of the model is updated (if parameter values have changed),
while getLogProb() simply returns the cached value.

```{r}
lmModel$b0
lmModel$b0 <- 2

lmModel$getLogProb()
lmModel$calculate()
lmModel$getLogProb()
```

Another useful method is getDependencies(), which returns all nodes that are 
dependent on the input argument.

```{r}
lmModel$getDependencies('b0')
```

We can combine getDependencies() and getLogProb() to easily find the full 
conditional of a specified parameter.

```{r}
depNodes <- lmModel$getDependencies('b0')
lmModel$getLogProb(depNodes)

sum(dnorm(lmModel$y, mean = lmModel$b0 + lmModel$b1*x, sd = lmModel$sigma, log = T)) +
    dnorm(lmModel$b0, 0, 100, log = T)
```


# Writing a custom sampler

We can use these methods to write a custom sampler function.

For example, we could create an independence sampler for sigma, where each proposed 
value is a random draw from the prior a Uniform(0, 100) distribution.

```{r}
indSampler <- nimbleFunction(
    name = 'indSampler',                         # Convenient for class name of R reference class and generated C++ class
    contains = sampler_BASE,                     # There is a simple class inheritance system.
    setup = function(model, mvSaved, target, control) {                 # REQUIRED setup arguments
        scale <- if(!is.null(control$scale)) control$scale else 1         # Typical extraction of control choices
        calcNodes <- model$getDependencies(target)                        # Typical query of model structure
    },                                                                  # setup can't return anything
    run = function() {
        currentValue <- model[[target]]                                   # extract current value
        currentLogProb <- model$getLogProb(calcNodes)                     # get log "denominator" from cached values
        
        proposalValue <- runif(1, 0, 100)        # generate proposal value
        
        model[[target]] <<- proposalValue                                 # put proposal value in model
        proposalLogProb <- model$calculate(calcNodes)                     # calculate log "numerator" 
        logAcceptanceRatio <- proposalLogProb - currentLogProb            # log acceptance ratio
        
        accept <- decide(logAcceptanceRatio)                              # utility function to generate accept/reject decision
        if(accept)                                                        # accept: synchronize model -> mvSaved
            copy(from = model, to = mvSaved, row = 1, nodes = calcNodes, logProb = TRUE)
        else                                                              # reject: synchronize mvSaved -> model
            copy(from = mvSaved, to = model, row = 1, nodes = calcNodes, logProb = TRUE)
    },
    methods = list(                              # required method for sampler_BASE base class
        reset = function() {}
    )
)
```


Then we can modify our MCMC configuration to use this sampler

```{r}
lmConfigure$removeSamplers('sigma')
lmConfigure$addSampler('sigma',
                       type = 'indSampler')

lmConfigure

lmMCMC <- buildMCMC(lmConfigure)

lmCompiled <- compileNimble(lmModel, lmMCMC) 

samples4 <- runMCMC(lmCompiled$lmMCMC, 
                    niter = 100000,
                    nburnin = 10000)

head(samples4)

par(mfrow = c(1,3))
plot(samples4[,'b0'], type = 'l')
abline(h = b0_true, col = 'red')
plot(samples4[,'b1'], type = 'l')
abline(h = b1_true, col = 'red')
plot(samples4[,'sigma'], type = 'l')
abline(h = sigma_true, col = 'red')
```


This is a very inefficient sampler.

# Simulating from a NIMBLE model

We can also use nimble models to simulate data. This is useful for simulation
studies and posterior prediction.

```{r}
### create nimble model object without data or initial values
lmModelSim <- nimbleModel(lmCode, 
                          constants = constantsList)

# exclude data from parent nodes
parentNodes <- lmModelSim$getParents('y', stochOnly = TRUE)
nodesToSim <- lmModelSim$getDependencies(parentNodes, self = FALSE,downstream = T)

lmModelSim$b0 <- b0_true
lmModelSim$b1 <- b1_true
lmModelSim$sigma <- sigma_true

nSim <- 500

system.time({
    simDat <- matrix(NA, nrow = constantsList$n, ncol = nSim)
    for (i in 1:nSim) {
        lmModelSim$simulate(nodesToSim, includeData = TRUE)
        simDat[,i] <- lmModelSim$y
    }
})


par(mfrow = c(2,2)) 
for (i in 1:4) {
    plot(constantsList$x, simDat[,i])
}


```

Writing all of this into a function which can be compiled in C++ yields incredibly
fast run times.

```{r}
# function to simulate from model
simulator <- nimbleFunction(
  setup = function(model, dataNodes) {
      
    parentNodes <- model$getParents(dataNodes, stochOnly = TRUE)
    nodesToSim <- model$getDependencies(parentNodes, self = FALSE,downstream = T)

    nData <- length(model$expandNodeNames(dataNodes, returnScalarComponents = TRUE))
  },
  run = function(nSim = double(0)) {
    simDat <- matrix(nrow = nData, ncol = nSim)   
    for(i in 1:nSim) {
      model$simulate(nodesToSim, includeData = TRUE)
      simDat[, i] <- values(model, dataNodes)
    }
    return(simDat)
    returnType(double(2))
  })


compiled <- compileNimble(lmModelSim)

sim_R <- simulator(model= lmModelSim, dataNodes = 'y')
sim_C <- compileNimble(sim_R)

nSim <- 10000

system.time({
   simDat <- sim_C$run(nSim)
})

par(mfrow = c(2,2)) 
for (i in 1:4) {
    plot(constantsList$x, simDat[,i])
}

```







