---
title: "Bayesian SIR Model - Incomplete Data"
output: slidy_presentation
---

# Only Incidence is Observed 

```{r}
library(nimble)
```

# Model Code

Model code is exactly the same as before. Differences will arise in specifying the 
data when fitting the model.

```{r}
SIR_code <-  nimbleCode({
    
    S[1] <- N - I0 - R0
    I[1] <- I0
    R[1] <- R0
    
    probIR <- 1 - exp(-gamma)
    
    ### loop over time
    for(t in 1:tau) {
        
        probSI[t] <- 1 - exp(- beta * I[t] / N)
        
        Istar[t] ~ dbin(probSI[t], S[t])
        Rstar[t] ~ dbin(probIR, I[t])
        
        # update S, I, R
        S[t + 1] <- S[t] - Istar[t]
        I[t + 1] <- I[t] + Istar[t] - Rstar[t]
        R[t + 1] <- R[t] + Rstar[t] 
        
    }
    
    # priors
    beta ~ dgamma(0.1, 0.1)
    gamma ~ dgamma(aa, bb)
    
})
```


# Simulate an example epidemic to use for model fitting

```{r}
constantsList <- list(N = 10000,
                      I0 = 5,
                      R0 = 0,
                      tau = 60)


sirModel <- nimbleModel(SIR_code, 
                        constants = constantsList)

# exclude data from parent nodes
dataNodes <- c('Istar', 'Rstar')
dataNodes <- sirModel$expandNodeNames(dataNodes, returnScalarComponents = TRUE)
parentNodes <- sirModel$getParents(dataNodes, stochOnly = TRUE)
parentNodes <- parentNodes[-which(parentNodes %in% dataNodes)]
parentNodes <- sirModel$expandNodeNames(parentNodes, returnScalarComponents = TRUE)
nodesToSim <- sirModel$getDependencies(parentNodes, self = FALSE, downstream = T)

initsList <- list(beta = 0.6,
                  gamma = 0.2)
sirModel$setInits(initsList)

set.seed(1)
sirModel$simulate(nodesToSim, includeData = TRUE)

trueIstar <- sirModel$Istar

plot(trueIstar, main = 'Epidemic curve', xlab = 'Time', ylab = 'Incidence', 
     type = 'l')

```


# Writing the Proposal Function

A proposal function in NIMBLE has some restrictions in formatting/functions that 
can be used as it will be compiled in C++.

https://r-nimble.org/html_manual/cha-RCfunctions.html#sec:r-fiunctions-implemented 

```{r}
RstarUpdate <- nimbleFunction(
    name = 'RstarUpdate',                              
    contains = sampler_BASE,                     
    setup = function(model, mvSaved, target, control) {  # REQUIRED setup arguments
        calcNodes <- model$getDependencies(target) 
        
        # number of update attempts at each iteration
        nUpdates <- 100
    },        # setup can't return anything
    run = function() {
        currentValue <- model[[target]]                                   
        currentLogProb <- model$getLogProb(calcNodes)     
        
        nTimePoints <- length(currentValue)
        
        # repeat proposal many times 
        for (it in 1:nUpdates) {
            
            # three possible moves:
            moveType <- ceiling(runif(1, 0, 3))
            proposalValue <- currentValue
            
            if (moveType == 1) {
                # add a removal time
                addIdx <- runif(1, 1, nTimePoints + 1)
                proposalValue[addIdx] <- proposalValue[addIdx] + 1
                
                # g(old|new) - g(new|old)
                # subtract from new - add to old
                possibleSubtract <- which(proposalValue > 0)
                g <- -log(length(possibleSubtract)) + log(nTimePoints)
                
                
            } else if (moveType == 2) {
                # move a removal time
                possibleSubtract <- which(currentValue > 0)
                subtractIdx <- possibleSubtract[runif(1, 1, length(possibleSubtract) + 1)]
                addIdx <- runif(1, 1, nTimePoints + 1)
                
                proposalValue[subtractIdx] <- proposalValue[subtractIdx] - 1
                proposalValue[addIdx] <- proposalValue[addIdx] + 1
                
                # g(old|new) - g(new|old)
                # possibly have different number of values to subtract from 
                newPossibleSubtract <- which(proposalValue > 0)
                g <- -log(length(newPossibleSubtract)) +log(length(possibleSubtract))
                
            } else if (moveType == 3) {
                # subtract a removal time
                possibleSubtract <- which(currentValue > 0)
                subtractIdx <- possibleSubtract[runif(1, 1, length(possibleSubtract) + 1)]
                proposalValue[subtractIdx] <- proposalValue[subtractIdx] - 1
                
                # g(old|new) - g(new|old)
                # add to new - subtract from old
                g <- -log(nTimePoints) + log(length(possibleSubtract)) 
                
            }
            
            # put proposal value in model
            model[[target]] <<- proposalValue                                
            proposalLogProb <- model$calculate(calcNodes)                     
            logAcceptanceRatio <- proposalLogProb - currentLogProb + g            
            
            accept <- decide(logAcceptanceRatio)                              
            
            if (accept) {
                # no changes to model object needed
                currentLogProb <- proposalLogProb
                currentValue <- proposalValue
                
            } else {
                # reject proposal and revert model to current state
                model[[target]] <<- currentValue
                
                # current full conditional (calculate overwrites the stored value)
                currentLogProb <- model$calculate(calcNodes) 
            }
            
        } # end loop
        
        # synchronize model -> mvSaved after nUpdates
        copy(from = model, to = mvSaved, row = 1, nodes = calcNodes, logProb = TRUE)
        
    },
    methods = list(                              # required method for sampler_BASE base class
        reset = function() {}
    )
)
```


# Incomplete Data - Model Configuration and Sampling

$R^*$ is initialized by specifying all removals happen five days after infection.

```{r}
dataList <- list(Istar = trueIstar)

bb <- 348
aa <- 0.2 * bb

tau <- length(dataList$Istar)

constantsList <- list(N = 10000,
                      I0 = 5,
                      R0 = 0,
                      tau = tau,
                      aa = aa,
                      bb = bb) 

set.seed(4)
initsList <- list(beta = runif(1, 0, 1),
                  gamma = rgamma(1, aa, bb), 
                  Rstar = c(rep(0, 4),
                            constantsList$I0, 
                            head(dataList$Istar, -5)))


sirModelFit <- nimbleModel(SIR_code, 
                           constants = constantsList,
                           data = dataList,
                           inits = initsList)

with(sirModelFit, cbind(S=S[-(tau + 1)], Istar, I=I[-(tau + 1)], 
                        Rstar, R=R[-(tau + 1)]))[1:10,]



myConfig <- configureMCMC(sirModelFit)

myConfig$removeSamplers('Rstar') 
myConfig$addSampler(target = c('Rstar'),
                    type = "RstarUpdate")
myConfig$addMonitors(c('Rstar'))
myConfig

nimbleOptions(MCMCusePredictiveDependenciesInCalculations = TRUE)
myMCMC <- buildMCMC(myConfig)

system.time({
    compiled <- compileNimble(sirModelFit, myMCMC) 
    samples <- runMCMC(compiled$myMCMC, niter = 100000, nburnin = 50000, setSeed = 5)
})

head(samples)
```

# Trace Plots of Parameters

```{r}
par(mfrow = c(1,2))
plot(samples[,'beta'], type = 'l')
abline(h = 0.6, col = 'red')
plot(samples[,'gamma'], type = 'l')
abline(h = 0.2, col = 'red')
```

# Posterior Distribution of Estimated Removal Times

```{r}
RstarSamples <- samples[,grep('Rstar', colnames(samples))]

RstarMean <- colMeans(RstarSamples)
RstarCI <- apply(RstarSamples, 2, quantile, probs = c(0.025, 0.975))

plot(RstarMean, type = 'l', ylim = c(0, 700))
lines(RstarCI[1,], lty = 2)
lines(RstarCI[2,], lty = 2)
lines(sirModel$Rstar, col = 'red')
legend('topright', c('posterior mean', 'truth'), col = c('black', 'red'), lwd = 1)
```


# Evaluating Prior Sensitivity for Gamma

The previously used prior puts 90% probability on the mean infectious period 
being between 4 and 6 days, centered on 5 days.

```{r}
bb <- 348
aa <- 0.2 * bb
pgamma(1/4, aa, bb) - pgamma(1/6, aa, bb)
curve(dgamma(x, aa, bb))
```

Now, we will compare this with a weaker and stronger prior, both also centered
on 5 days.

```{r, fig.width = 12}
par(mfrow = c(1,2))
bb_2 <- 57
aa_2 <- 0.2 * bb_2
pgamma(1/4, aa_2, bb_2) - pgamma(1/6, aa_2, bb_2)
curve(dgamma(x, aa_2, bb_2), main = 'weaker prior')


bb_3 <- 2000
aa_3 <- 0.2 * bb_3
pgamma(1/4, aa_3, bb_3) - pgamma(1/6, aa_3, bb_3)
curve(dgamma(x, aa_3, bb_3), main = 'stronger prior')
```

We will also use a joint MH proposal for beta and gamma to try and account for the correlation 
between these two parameters.

```{r}
dataList <- list(Istar = trueIstar)

# Weaker prior

constantsList <- list(N = 10000,
                      I0 = 5,
                      R0 = 0,
                      tau = length(dataList$Istar),
                      aa = aa_2,
                      bb = bb_2) 

set.seed(8)
initsList <- list(beta = runif(1, 0, 1),
                  gamma = rgamma(1, aa_2, bb_2), 
                  Rstar = c(rep(0, 4),
                            constantsList$I0, 
                            head(dataList$Istar, -5)))


sirModelFit <- nimbleModel(SIR_code, 
                           constants = constantsList,
                           data = dataList,
                           inits = initsList)

myConfig <- configureMCMC(sirModelFit)

myConfig$removeSamplers('Rstar') 
myConfig$addSampler(target = c('Rstar'),
                    type = "RstarUpdate")
myConfig$addMonitors(c('Rstar'))

# block sampler for transmission parameters
paramsForBlock <- c('beta', 'gamma')
myConfig$removeSampler(paramsForBlock)
myConfig$addSampler(target = paramsForBlock, type = "RW_block",
                    control = list(adaptInterval = 100,
                                   propCov = diag(c(0.2, 0.2))))

myConfig

nimbleOptions(MCMCusePredictiveDependenciesInCalculations = TRUE)
myMCMC <- buildMCMC(myConfig)

system.time({
    compiled <- compileNimble(sirModelFit, myMCMC) 
    samplesWeak <- runMCMC(compiled$myMCMC, niter = 100000, nburnin = 50000, setSeed = 15)
})


#### Stronger prior

dataList <- list(Istar = trueIstar)

constantsList <- list(N = 10000,
                      I0 = 5,
                      R0 = 0,
                      tau = length(dataList$Istar),
                      aa = aa_3,
                      bb = bb_3) 

set.seed(1)
initsList <- list(beta = runif(1, 0, 1),
                  gamma = rgamma(1, aa_3, bb_3), 
                  Rstar = c(rep(0, 4),
                            constantsList$I0, 
                            head(dataList$Istar, -5)))


sirModelFit <- nimbleModel(SIR_code, 
                           constants = constantsList,
                           data = dataList,
                           inits = initsList)

myConfig <- configureMCMC(sirModelFit)

myConfig$removeSamplers('Rstar') 
myConfig$addSampler(target = c('Rstar'),
                    type = "RstarUpdate")
myConfig$addMonitors(c('Rstar'))

# block sampler for transmission parameters
paramsForBlock <- c('beta', 'gamma')
myConfig$removeSampler(paramsForBlock)
myConfig$addSampler(target = paramsForBlock, type = "RW_block",
                    control = list(adaptInterval = 100,
                                   propCov = diag(c(0.2, 0.2))))

myConfig

nimbleOptions(MCMCusePredictiveDependenciesInCalculations = TRUE)
myMCMC <- buildMCMC(myConfig)

system.time({
    compiled <- compileNimble(sirModelFit, myMCMC) 
    samplesStrong <- runMCMC(compiled$myMCMC, niter = 100000, nburnin = 50000, setSeed = 12)
})


```


```{r, fig.width = 12, fig.height=12}
par(mfrow = c(2,2))
plot(samplesWeak[,'beta'], type = 'l', main = 'beta - weaker prior')
abline(h = 0.6, col = 'red')
plot(samplesWeak[,'gamma'], type = 'l', main = 'gamma - weaker prior')
abline(h = 0.2, col = 'red')
plot(samplesStrong[,'beta'], type = 'l', main = 'beta - stronger prior')
abline(h = 0.6, col = 'red')
plot(samplesStrong[,'gamma'], type = 'l', main = 'gamma - stronger prior')
abline(h = 0.2, col = 'red')
```

# Compare Posterior distributions

```{r, fig.width = 12}
par(mfrow = c(1,3))
hist(samples[,'gamma'], freq = F, main = 'initial prior', xlim = c(0.12, 0.26))
curve(dgamma(x, aa, bb), add = T, col = 'red')

hist(samplesWeak[,'gamma'], freq = F, main =  'weaker prior', xlim = c(0.12, 0.26))
curve(dgamma(x, aa_2, bb_2), add = T, col = 'red')

hist(samplesStrong[,'gamma'], freq = F, main =  'stronger prior', xlim = c(0.12, 0.26))
curve(dgamma(x, aa_3, bb_3), add = T, col = 'red')

summary(samples[,'gamma'])
summary(samplesWeak[,'gamma'])
summary(samplesStrong[,'gamma'])
```

