---
title: "Simulation from a  Spatial ILM <br>  (SIR and binary covariate)"
output: slidy_presentation
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = FALSE)
```


#  Spatial ILM (SIR and binary covariate)

```{r}
library(EpiILM)
```




# Simulation Spatial Population

```{r, include=TRUE}
 
x = runif(100, 0, 10)
y = runif(100, 0, 10)
```

We will artificially put all high biosecurity farms in western half of population, all low  biosecurity farms in east (for visual effect!)
```{r, include=TRUE}
bio = rep(0,100)
bio[x >= 5] = 1
plot (x,y, cex=0.6, pch=bio*2)
```

# Simulation Epidemic Through Population

```{r, include=TRUE}
infectious = rep(3,100)
SIR.dis = epidata(type = "SIR", n = 100, tmax = 15, Sformula = ~ bio, sus.par = c(0.3, 0.6), 
                 beta = 5.0, infperiod = infectious, x = x, y = y)
```


```{r}
SIR.dis
```



# Epidemic Plots 


```{r}
plot(SIR.dis, plottype = "curve", curvetype ="complete")
```

# Epidemic Plots 
```{r}
plot(SIR.dis, plottype = "spatial")
```



# Multiple Epidemics

Let’s repeat the epidemic simulation 20 times 

Choose initial infection to be individual 10 at time point 3

```{r}
inftime <- rep(0,100)
inftime[10] <- 3

out1 <- vector(mode = "list", length = 20)
for (i in 1:20){
    dat <- epidata(type = "SIR", n = 100, tmax = 50, Sformula = ~ bio, sus.par = c(0.3, 0.6), 
                 beta = 5.0, infperiod = infectious, x = x, y = y, inftime = inftime)
    out1[[i]] <- dat$inftime }
```




Calculate the number of new infections at each time point
```{r}
infect1 <- vector(mode = "list", length = 20)
for (j in 1:20){
    newinf <- rep(0)
     for (i in 1:50){
     newinf[i] <- length(out1[[j]][out1[[j]]==i]) }
     infect1[[j]] <- newinf }
```

#

Plot this:

```{r}
infectmax <- vector(mode = "list", length = 20)
  for (j in 1:20){ infectmax[[j]]=max(infect1[[j]] ) }
  infectmax <- unlist(infectmax)
  max1 <- max(infectmax)

  plot(seq(1:50),infect1[[1]],type="b",pch=20,lwd=2,
    xaxt="n",xlab="time ",ylab="New infections",ylim=c(0,max1))
  axis(1, at=1:50)
  for (j in 2:20){
    lines(seq(1:50),infect1[[j]],type="b",pch=20,lwd=2,
    xaxt="n",xlab="",yaxt="n",ylab="") }
  
```
  

