---
title: "SEIR Model with Intervention"
output: slidy_presentation
---

# SEIR Model Example - Ebola

```{r}
library(nimble)
library(outbreaks)
library(ggplot2)
library(plyr)
```

# 2014-2015 Ebola in Sierra Leone


```{r}
ebola <- ebola_sierraleone_2014
ebola <- ddply(ebola, .(date_of_onset), summarize, 
               cases = length(id))

# fill in missing dates with 0 cases
allDates <- data.frame(date_of_onset = seq(min(ebola$date_of_onset),
                                           max(ebola$date_of_onset), 1))
ebola <- merge(allDates, ebola, by = 'date_of_onset', all.x = T)
ebola <- ebola[order(ebola$date_of_onset),]
ebola$day <- 1:nrow(ebola)
ebola$cases[is.na(ebola$cases)] <- 0

head(ebola)

summary(ebola[,-1])

plot(ebola$date_of_onset, ebola$cases, type = 'h', main = 'Incidence')
abline(v = as.Date('2014-10-01'), col = 'red')
abline(v = as.Date('2014-12-28'), col = 'red')

```



# Writing the SEIR model

To incorporate the intervention effects, we allow transmission to be piecewise
constant.

\[ \beta_t = \exp (\boldsymbol{X} \boldsymbol{\beta}) \]

```{r}

SEIR_code <-  nimbleCode({
  
  S[1] <- N - E0 - I0 - R0
  E[1] <- E0
  I[1] <- I0
  R[1] <- R0
  
  probEI <- 1 - exp(-rho)
  probIR <- 1 - exp(-gamma)
  
  
  ### loop over time
  for(t in 1:tau) {
    
    betat[t] <- inprod(beta[1:3], X[t, 1:3])
    
    probSE[t] <- 1 - exp(- exp(betat[t]) * I[t] / N)
    
    Estar[t] ~ dbin(probSE[t], S[t])
    Istar[t] ~ dbin(probEI, E[t])
    Rstar[t] ~ dbin(probIR, I[t])
    
    # update S, I, R
    S[t + 1] <- S[t] - Estar[t]
    E[t + 1] <- E[t] + Estar[t] - Istar[t]
    I[t + 1] <- I[t] + Istar[t] - Rstar[t]
    R[t + 1] <- R[t] + Rstar[t] 
    
    # compute R0(t)
    reproNum[t] <- exp(betat[t]) / gamma
    
  }
  
  # priors
  for (i in 1:3) {
    beta[i] ~ dnorm(0, sd = 4)
  }
  gamma ~ dgamma(aa, bb)
  rho ~ dgamma(cc, dd)
  
})
```


# Specifying priors

Ebola has a latent/incubation period which is 8 to 10 days on average. 

```{r}
dd <- 5000
cc <- 1/9 * dd
pgamma(1/8, cc, dd) - pgamma(1/10, cc, dd)
curve(dgamma(1/x, cc, dd), from = 5, to = 14, 
      main = 'prior for mean latent period')
```

Individuals can only transmit EVD to others after symptoms have appeared, and remain infectious for around 4 to 10 days.

```{r}
bb <- 2000
aa <- 1/7 * bb
pgamma(1/6, aa, bb) - pgamma(1/8, aa, bb)
curve(dgamma(1/x, aa, bb), from = 1, to = 14, 
      main = 'prior for mean infectious period')

```

# User-defined proposal for transitions

Proposal function for unobserved transition vectors $E^*$ and $R^*$

```{r}
transUpdate <- nimbleFunction(
  name = 'transUpdate',                              
  contains = sampler_BASE,                     
  setup = function(model, mvSaved, target, control) { # REQUIRED setup arguments
    calcNodes <- model$getDependencies(target) 
    
    # number of update attempts 
    nUpdates <- 500
  },  # setup can't return anything
  run = function() {
    currentValue <- model[[target]]                                   
    currentLogProb <- model$getLogProb(calcNodes)  
    
    nTimePoints <- length(currentValue)
    
    # repeat proposal many times 
    for (it in 1:nUpdates) {
      
      # three possible moves:
      moveType <- ceiling(runif(1, 0, 3))
      
      proposalValue <- currentValue
      
      if (moveType == 1) {
        # add a removal time
        addIdx <- runif(1, 1, nTimePoints + 1)
        proposalValue[addIdx] <- proposalValue[addIdx] + 1
        
        # g(old|new) - g(new|old)
        # subtract from new - add to old
        possibleSubtract <- which(proposalValue > 0)
        g <- -log(length(possibleSubtract)) + log(nTimePoints)
        
        
      } else if (moveType == 2) {
        # move a removal time
        possibleSubtract <- which(currentValue > 0)
        subtractIdx <- possibleSubtract[runif(1, 
                                              1, length(possibleSubtract) + 1)]
        addIdx <- runif(1, 1, nTimePoints + 1)
        
        proposalValue[subtractIdx] <- proposalValue[subtractIdx] - 1
        proposalValue[addIdx] <- proposalValue[addIdx] + 1
        
        # g(old|new) - g(new|old)
        # possibly have different number of values to subtract from 
        newPossibleSubtract <- which(proposalValue > 0)
        g <- -log(length(newPossibleSubtract)) +log(length(possibleSubtract))
        
      } else if (moveType == 3) {
        # subtract a removal time
        possibleSubtract <- which(currentValue > 0)
        subtractIdx <- possibleSubtract[runif(1, 
                                              1, length(possibleSubtract) + 1)]
        proposalValue[subtractIdx] <- proposalValue[subtractIdx] - 1
        
        # g(old|new) - g(new|old)
        # add to new - subtract from old
        g <- -log(nTimePoints) + log(length(possibleSubtract)) 
        
      }
      
      # put proposal value in model
      model[[target]] <<- proposalValue                                
      proposalLogProb <- model$calculate(calcNodes)                     
      logAcceptanceRatio <- proposalLogProb - currentLogProb + g            
      
      accept <- decide(logAcceptanceRatio)                              
      
      if (accept) {
        # no changes to model object needed
        currentLogProb <- proposalLogProb
        currentValue <- proposalValue
        
      } else {
        # reject proposal and revert model to current state
        model[[target]] <<- currentValue
        
        # current full conditional (calculate overwrites the stored value)
        currentLogProb <- model$calculate(calcNodes) 
      }
      
    } # end loop
    
    # synchronize model -> mvSaved after nUpdates
    copy(from = model, to = mvSaved, row = 1, nodes = calcNodes, logProb = TRUE)
    
  },
  methods = list(    # required method for sampler_BASE base class
    reset = function() {}
  )
)
```


# Model Fitting

```{r}
dataList <- list(Istar = ebola$cases[-1])


tau <- length(dataList$Istar)
tstar1 <- which(ebola$date == as.Date('2014-10-01')) - 1
tstar2 <- which(ebola$date == as.Date('2014-12-28')) - 1
X <- cbind(1, 1:tau >= tstar1, 1:tau >= tstar2)

constantsList <- list(N = 7092113,
                      E0 = sum(dataList$Istar[c(1:9)]),
                      I0 = ebola$cases[1],
                      R0 = 0,
                      tau = tau,
                      X = X,
                      aa = aa,
                      bb = bb,
                      cc = cc,
                      dd = dd) 

set.seed(2)
# initially assume exposed 9 days before infection and removed 7 days after
initsList <- list(beta = rnorm(3, 0, 2),
                  gamma = rgamma(1, aa, bb),
                  rho = rgamma(1, cc, dd),
                  Estar = c(dataList$Istar[-c(1:9)], 
                            rep(0, 9)), 
                  Rstar = c(rep(0, 6), constantsList$I0,
                            dataList$Istar[1:(tau-7)]))


seirModel <- nimbleModel(SEIR_code, 
                         constants = constantsList,
                         data = dataList,
                         inits = initsList)


myConfig <- configureMCMC(seirModel)

myConfig$removeSamplers('Estar') # Nodes will be expanded
myConfig$addSampler(target = c('Estar'),
                    type = "transUpdate")
myConfig$addMonitors(c('Estar'))

myConfig$removeSamplers('Rstar') # Nodes will be expanded
myConfig$addSampler(target = c('Rstar'),
                    type = "transUpdate")
myConfig$addMonitors(c('Rstar'))


myConfig$addMonitors(c('reproNum'))

myConfig


nimbleOptions(MCMCusePredictiveDependenciesInCalculations = TRUE)
myMCMC <- buildMCMC(myConfig)
compiled <- compileNimble(seirModel, myMCMC) 

```

```{r, eval = F}
samples <- runMCMC(compiled$myMCMC, niter = 60000, nburnin = 10000, setSeed = 1)
```


```{r, echo = F}
#saveRDS(samples, './DATA/ebolaSamples.rds')
samples <- readRDS('./DATA/ebolaSamples.rds')
```

# Trace plots for model parameters

```{r}
par(mfrow = c(2,3))
plot(samples[,'beta[1]'], type = 'l')
plot(samples[,'beta[2]'], type = 'l')
plot(samples[,'beta[3]'], type = 'l')
plot(samples[,'rho'], type = 'l')
plot(samples[,'gamma'], type = 'l')
```

# Posterior distribution of exposure and removal times

```{r}
EstarSamples <- samples[,grep('Estar', colnames(samples))]
EstarMean <- colMeans(EstarSamples)
EstarCI <- apply(EstarSamples, 2, quantile, probs = c(0.025, 0.975))

RstarSamples <- samples[,grep('Rstar', colnames(samples))]
RstarMean <- colMeans(RstarSamples)
RstarCI <- apply(RstarSamples, 2, quantile, probs = c(0.025, 0.975))

transSummary <- data.frame(date = ebola$date_of_onset[-1], 
                         meanE = EstarMean,
                         lowerE = EstarCI[1,],
                         upperE = EstarCI[2,], 
                         meanR = RstarMean,
                         lowerR = RstarCI[1,],
                         upperR = RstarCI[2,])

ggplot(transSummary, aes(x = date, y = meanE, ymin = lowerE, ymax = upperE)) +
  geom_line(color = 'blue', size = 1) + 
  geom_ribbon(fill = 'deepskyblue', alpha = 0.3) + 
  theme_bw() +
  ggtitle('Estimated exposure times')


ggplot(transSummary, aes(x = date, y = meanR, ymin = lowerR, ymax = upperR)) +
  geom_line(color = 'blue', size = 1) + 
  geom_ribbon(fill = 'deepskyblue', alpha = 0.3) + 
  theme_bw() +
  ggtitle('Estimated removal times')
```


# Posteriors for mean latent and infectious periods

Histogram of posterior compared to prior distribution (red)

```{r, fig.width = 10}
par(mfrow = c(1,2))
hist(1/samples[,'rho'], freq = F, xlim = c(4, 14))
x <- 1/rgamma(1e6, cc, dd)
dx <- density(x)
lines(dx, col = "red")

hist(1/samples[,'gamma'], freq = F, xlim = c(0, 14))
x <- 1/rgamma(1e6, aa, bb)
dx <- density(x)
lines(dx, col = "red")
```


# Reproductive number over time

Posterior mean and 95% credible intervals.

```{r}
r0Samples <- samples[,grep('reproNum', colnames(samples))]

r0Mean <- colMeans(r0Samples)
r0CI <- apply(r0Samples, 2, quantile, probs = c(0.025, 0.975))

r0Summary <- data.frame(date = ebola$date_of_onset[-1], 
                        mean = r0Mean,
                        lower = r0CI[1,],
                        upper = r0CI[2,])

ggplot(r0Summary, aes(x = date, y = mean, ymin = lower, ymax = upper)) +
  geom_line(color = 'blue', size = 1) + 
  geom_ribbon(fill = 'deepskyblue', alpha = 0.3) + 
  theme_bw() +
  geom_hline(yintercept = 1, linetype = 2) +
  ggtitle('Sierra Leone reproductive number over time')

```

Previous research found the reproductive number in Sierra Leone to be 

* around 1.2 or 1.3 https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4169299/ 

* 1.4 or 1.5 https://idpjournal.biomedcentral.com/articles/10.1186/s40249-015-0043-3#Sec9

* 1.70 (interquartile range: 1.34, 2.05) https://www.cambridge.org/core/journals/epidemiology-and-infection/article/systematic-review-of-early-modelling-studies-of-ebola-virus-disease-in-west-africa/154353B9A815326FE3656046AD6390B6


# Posterior Prediction to Assess Intervention Effect

```{r}
# function to simulate from model
simulator <- nimbleFunction(
  setup = function(model, dataNodes) {
    parentNodes <- model$getParents(dataNodes, stochOnly = TRUE)
    # exclude data from parent nodes
    parentNodes <- parentNodes[-which(parentNodes %in% dataNodes)]
    parentNodes <- model$expandNodeNames(parentNodes, 
                                         returnScalarComponents = TRUE)
    cat("Stochastic parents of data are: ", paste(parentNodes, sep = ','), ".\n")
    simNodes <- model$getDependencies(parentNodes, self = FALSE,
                                      downstream = T)
    
    nData <- length(model$expandNodeNames(dataNodes, 
                                          returnScalarComponents = TRUE))
  },
  run = function(params = double(1), nSim = double()) {
    simDat <- matrix(nrow = nSim, ncol = nData)   
    for(i in 1:nSim) {
      values(model, parentNodes) <<- params
      model$simulate(simNodes, includeData = TRUE)
      simDat[i, ] <- values(model, dataNodes)
    }
    return(simDat)
    returnType(double(2))
  })


# prediction range is post intervention 1
dataNodes <- c(paste0('Estar[', tstar1:tau, ']'),
               paste0('Istar[', tstar1:tau, ']'),
               paste0('Rstar[', tstar1:tau, ']'))

sim_R <- simulator(seirModel, dataNodes)
sim_C <- compileNimble(sim_R)

# get order of parameters
parentNodes <- seirModel$getParents(dataNodes, stochOnly = TRUE)
parentNodes <- parentNodes[-which(parentNodes %in% dataNodes)]
parentNodes <- seirModel$expandNodeNames(parentNodes, returnScalarComponents = TRUE)

obsIstar <- seirModel$Istar[1:(tstar1 - 1)]
names(obsIstar) <- paste0('Istar[', 1:(tstar1 - 1), ']')
```


If control measures were never introduced

```{r}
nPost <- 10000
postPredInc <- matrix(NA, nrow = length( tstar1:tau), ncol = nPost)
set.seed(1)
for (k in 1:nPost) {
  
  postIdx <- sample(1:nrow(samples), 1)
  
  EstarSample_post <- EstarSamples[postIdx,]
  RstarSample_post <- RstarSamples[postIdx,]
  paramsSample_post <- c(samples[postIdx, 'beta[1]'],
                         'beta[2]' = 0,
                         'beta[3]' = 0,
                         samples[postIdx,'rho'],
                         samples[postIdx,'gamma'])
  
  allSamples <- c(paramsSample_post, EstarSample_post, 
                  obsIstar, RstarSample_post)
  allSamples <- allSamples[parentNodes]
  
  tmp <- sim_C$run(params = allSamples, nSim = 10)
  predInc <- tmp[,grep('Istar', dataNodes)]
  postPredInc[,k] <- apply(predInc, 2, median)
}

```



```{r}
postPredMean <- rowMeans(postPredInc)
postPredCI <- apply(postPredInc, 1, quantile, probs = c(0.025, 0.975))

ebola$time <- 0:tau
postPredDat <- data.frame(time =tstar1:tau, 
                          mean = postPredMean,
                          lower = postPredCI[1,],
                          upper = postPredCI[2,])
postPredDat <- merge(ebola, postPredDat, by = 'time', all.x = T)


ggplot(postPredDat, aes(x = date_of_onset, y = mean, ymin = lower, ymax = upper)) +
  geom_line(col= 'red') + 
  geom_line(aes(y = cases)) + 
  geom_ribbon(fill = 'red', alpha = 0.3) + 
  theme_bw() +
  labs(y = 'New cases', x= 'Date') +
  xlim(as.Date('2014-05-18'),
       ebola$date_of_onset[which(ebola$time == tstar1 + 50)]) +
  ylim(0, 350)


ggplot(postPredDat, aes(x = date_of_onset, y = mean, ymin = lower, ymax = upper)) +
  geom_line(col= 'red') + 
  geom_line(aes(y = cases)) + 
  geom_ribbon(fill = 'red', alpha = 0.3) + 
  theme_bw() +
  labs(y = 'New cases', x= 'Date') 
```

```{r}
sum(postPredMean)
sum(dataList$Istar[tstar1:tau])

sum(postPredMean) - sum(dataList$Istar[tstar1:tau])
```

Without control measures, we would expect 3,215,735 more people would have 
contracted EVD during the study period

# Less Effective Control Measures

If control measures never achieved 100% case isolation and safe burials.

```{r}
# prediction range is post intervention 1
dataNodes <- c(paste0('Estar[', tstar2:tau, ']'),
               paste0('Istar[', tstar2:tau, ']'),
               paste0('Rstar[', tstar2:tau, ']'))

sim_R <- simulator(seirModel, dataNodes)
sim_C <- compileNimble(sim_R)

# get order of parameters
parentNodes <- seirModel$getParents(dataNodes, stochOnly = TRUE)
parentNodes <- parentNodes[-which(parentNodes %in% dataNodes)]
parentNodes <- seirModel$expandNodeNames(parentNodes, returnScalarComponents = TRUE)

obsIstar <- seirModel$Istar[1:(tstar2 - 1)]
names(obsIstar) <- paste0('Istar[', 1:(tstar2 - 1), ']')

nPost <- 10000
postPredInc <- matrix(NA, nrow = length( tstar2:tau), ncol = nPost)
set.seed(1)
for (k in 1:nPost) {
  
  postIdx <- sample(1:nrow(samples), 1)
  
  EstarSample_post <- EstarSamples[postIdx,]
  RstarSample_post <- RstarSamples[postIdx,]
  paramsSample_post <- c(samples[postIdx, 'beta[1]'],
                         samples[postIdx, 'beta[2]'],
                         'beta[3]' = 0,
                         samples[postIdx,'rho'],
                         samples[postIdx,'gamma'])
  
  allSamples <- c(paramsSample_post, EstarSample_post, 
                  obsIstar, RstarSample_post)
  allSamples <- allSamples[parentNodes]
  
  tmp <- sim_C$run(params = allSamples, nSim = 10)
  predInc <- tmp[,grep('Istar', dataNodes)]
  postPredInc[,k] <- apply(predInc, 2, median)
}

```


```{r}
postPredMean <- rowMeans(postPredInc)
postPredCI <- apply(postPredInc, 1, quantile, probs = c(0.025, 0.975))

postPredDat <- data.frame(time =tstar2:tau, 
                          mean = postPredMean,
                          lower = postPredCI[1,],
                          upper = postPredCI[2,])
postPredDat <- merge(ebola, postPredDat, by = 'time', all.x = T)

ggplot(postPredDat, aes(x = date_of_onset, y = mean, ymin = lower, ymax = upper)) +
  geom_line(col= 'red') + 
  geom_line(aes(y = cases)) + 
  geom_ribbon(fill = 'red', alpha = 0.3) + 
  theme_bw() +
  labs(y = 'New cases', x= 'Date') 

```

```{r}
sum(postPredMean)
sum(dataList$Istar[tstar2:tau])

sum(postPredMean) - sum(dataList$Istar[tstar2:tau])
```

Without achieving 100% case isolation and safe burials, we expect 
6148 more people would have contracted EVD during the study period.




