---
title: "SIR Model with Infectious Duration Dependent Transmissibility"
output: slidy_presentation
---

# Infectious Duration Dependent Transmissibility

```{r}
library(nimble)
library(splines)
library(ggplot2)
```

# Writing the model for simulation

Flat priors are used because values will be directly specified when simulating,
but providing a prior distribution flags the node as stochastic.

```{r}
# nimbleFunction for logistic decay
logitDecay <- nimbleFunction(     
    run = function(x = double(1), w0 = double(0), k = double(0)) {
        returnType(double(1))
        
        result <- 1 / (1 + exp(k * (x - w0)))
        
        return(result)
    })

# nimbleFunction to get transmission probability
p_SI_t <- nimbleFunction(     
    run = function(beta = double(0), I = double(1), 
                   idd_curve = double(1), N = double(0)) {
        returnType(double(0))
        
        transRate <-  beta * sum(I * idd_curve) / N
        result <- 1 - exp(-transRate)
        
        return(result)
    })

# model code for simulation
SIR_IDD_code <-  nimbleCode({
    
    S[1] <- N - sum(I0[1:maxInf]) - R0
    I[1, 1:maxInf] <- I0[1:maxInf]
    R[1] <- R0
    
    idd_curve[1:maxInf] <- logitDecay(1:maxInf, w0, k)
    
    ### loop over time
    for(t in 1:tau) {
        
        probSI[t] <- p_SI_t(beta[t], I[t, 1:maxInf], idd_curve[1:maxInf], N)
        
        Istar[t] ~ dbin(probSI[t], S[t])
        
        # update S, I, R
        S[t + 1] <- S[t] - Istar[t]
        I[t + 1, 2:maxInf] <- I[t, 1:(maxInf - 1)]  # shift current I by one day
        I[t + 1, 1] <- Istar[t]                     # add newly infectious
        R[t + 1] <- R[t] + I[t, maxInf]
        
        beta[t] ~ dflat()
    }
    
    # priors
    w0 ~ dflat()
    k ~ dflat()
    
})
```

# Specify transmission over epidemic time

```{r}
# length of epidemic
tau <- 150

# matrix to describe intensity process
x <- cumsum(1:tau > 60) / 100
beta1 <- -1.7
beta2 <- -4
betat <- exp(beta1 + beta2 * x)

plot(betat, main = 'beta[t]', xlab = 'Epidemic time')
```

# Specify the IDD Curve

```{r}

# length of infectious period
maxInf <- 15

# IDD logistic decay parameters
w0 <- 8
k <- 1.5


plot(logitDecay(1:maxInf, w0, k), type = 'l', xlab = 'Day of infection')
```

# Simulate an epidemic

```{r}

constantsList <- list(tau = tau,
                      N = 1e6,
                      I0 = c(2, rep(0, maxInf - 1)),
                      R0 = 0,
                      maxInf = maxInf)

initsList <- list(beta = betat,
                  w0 = w0,
                  k = k)


iddModel <- nimbleModel(SIR_IDD_code, 
                        constants = constantsList,
                        inits = initsList)



# exclude data from parent nodes
dataNodes <- c('Istar')
dataNodes <- iddModel$expandNodeNames(dataNodes, returnScalarComponents = TRUE)
parentNodes <- iddModel$getParents(dataNodes, stochOnly = TRUE)
parentNodes <- parentNodes[-which(parentNodes %in% dataNodes)]
parentNodes <- iddModel$expandNodeNames(parentNodes, returnScalarComponents = TRUE)
nodesToSim <- iddModel$getDependencies(parentNodes, self = FALSE, downstream = T)

set.seed(1)
iddModel$simulate(nodesToSim, includeData = TRUE)


iddModel$I0
head(iddModel$I, 20)
head(iddModel$Istar, 20)

simI <- iddModel$I
colnames(simI) <- paste0('I_', 1:maxInf)

head(cbind(S = iddModel$S, Istar = iddModel$Istar, simI, R = iddModel$R), 20)

plot(iddModel$Istar, type = 'l', xlab = 'Epidemic time', ylab = 'New cases')
```


# Application - NYC Covid-19 first wave

```{r}
### Read in data from GitHub
nyc <-read.csv("https://raw.githubusercontent.com/nychealth/coronavirus-data/master/trends/cases-by-day.csv") 

nyc <- nyc[,c('date_of_interest', 'CASE_COUNT')]
colnames(nyc) <- c('date', 'dailyCases')

# format dates
nyc$date <- as.Date(nyc$date, format = '%m/%d/%Y')

# 7-day moving average of cases to account for reporting delays
# calculate moving average for smoothing
movingAverage <- function(x, bw) {
    n <- length(x)
    bw <- floor(bw)
    out <- rep(0, n)
    for (i in 1:n) {
        if (i < bw) {
            t1 = 1; t2 = i
        } else {
            t1 = i - bw + 1; t2 = i
        }
        out[i] <- mean(x[t1:t2])
    }
    return(out)
}
nyc$smoothedCases <- round(movingAverage(nyc$dailyCases, 7))

# population
nyc$Population <- 8.419*1e6

# peak 1 - Feb 29 - July 1, 2020
startDate <- as.Date('2020-02-29')
endDate <- as.Date('2020-07-01')
nycPeak1 <- nyc[nyc$date >= startDate & nyc$date < endDate,]

head(nycPeak1)

plot(nycPeak1$date, nycPeak1$smoothedCases, type = 'l')

```


# Model fitting

Use basis splines to model $\beta_t$.

```{r}

# nimbleFunction for logistic decay
logitDecay <- nimbleFunction(     
    run = function(x = double(1), w0 = double(0), k = double(0)) {
        returnType(double(1))
        
        result <- 1 / (1 + exp(k * (x - w0)))
        
        return(result)
    })

# nimbleFunction to get transmission probability
p_SI_t <- nimbleFunction(     
    run = function(beta = double(0), I = double(1), 
                   idd_curve = double(1), N = double(0)) {
        returnType(double(0))
        
        transRate <-  beta * sum(I * idd_curve) / N
        result <- 1 - exp(-transRate)
        
        return(result)
    })

# nimbleFunction to get R0(t)
getR0 <- nimbleFunction(     
    run = function(beta = double(0), idd_curve = double(1), N = double(0)) {
        returnType(double(0))
        
        result <- N * sum(1 - exp(- beta * idd_curve / N))
        
        return(result)
    })

nyc_code <-  nimbleCode({
    
    S[1] <- N - sum(I0[1:maxInf]) - R0
    I[1, 1:maxInf] <- I0[1:maxInf]
    R[1] <- R0
    
    idd_curve[1:maxInf] <- logitDecay(1:maxInf, w0, k)
    
    ### loop over time
    for(t in 1:tau) {
        
        beta[t] <- exp(inprod(b[1:p], XBasis[t, 1:p]))
        
        probSI[t] <- p_SI_t(beta[t], I[t, 1:maxInf], idd_curve[1:maxInf], N)
        
        Istar[t] ~ dbin(probSI[t], S[t])
        
        # update S, I, R
        S[t + 1] <- S[t] - Istar[t]
        I[t + 1, 2:maxInf] <- I[t, 1:(maxInf - 1)]  # shift current I by one day
        I[t + 1, 1] <- Istar[t]                     # add newly infectious
        R[t + 1] <- R[t] + I[t, maxInf]
        
        reproNum[t] <- getR0(beta[t],  idd_curve[1:maxInf], N)
        
        
    }
    
    # priors
    for (i in 1:p) {
        b[i] ~ dnorm(0, sd = 4)
    }
    w0 ~ dnorm(4, sd = 0.1)
    k ~ dgamma(100, 100)
    
})
```

Knots of the basis splines will be determined by important dates:

* March 7 - state of emergency declared

* March 16/17 - NYC public schools/bars and restaurants close

* March 22 - Stay at home order for non-essential workers

* June 8 - Phase 1 reopening

* June 22 - Phase 2 reopening



```{r}
# get data
incData <- nycPeak1$smoothedCases

# initial conditions from first 1 day
idxStart <- 1
I0 <- sum(incData[idxStart])

incData <- incData[-idxStart]

# length of infectious period
maxInf <- 10

# length of epidemic
tau <- length(incData)

# basis matrix (move left boundary so start point is not fixed at 0)

# knots at important change points
date1 <- which(nycPeak1$date[-idxStart] == as.Date('2020-03-07'))
date2 <- which(nycPeak1$date[-idxStart] == as.Date('2020-03-16'))
date3 <- which(nycPeak1$date[-idxStart] == as.Date('2020-03-22'))
date4 <- which(nycPeak1$date[-idxStart] == as.Date('2020-06-08'))
date5 <- which(nycPeak1$date[-idxStart] == as.Date('2020-06-22'))

knots <- c(date1, date2, date3, date4, date5)

XBasis <- bs(1:tau, Boundary.knots = c(-50, tau), knots = knots)
p <- ncol(XBasis)

constantsList <- list(tau = tau,
                      N = nycPeak1$Population[1],
                      I0 = c(I0, rep(0, maxInf - 1)),
                      R0 = 0,
                      maxInf = maxInf,
                      XBasis = XBasis,
                      p = p)

dataList <- list(Istar = incData)

nycModel <- nimbleModel(nyc_code, 
                        constants = constantsList,
                        data = dataList)

nycI <- nycModel$I
colnames(nycI) <- paste0('I_', 1:maxInf)
head(cbind(S = nycModel$S, Istar = nycModel$Istar, nycI, R = nycModel$R), 20)
```

Initial values are not specified so that they are drawn randomly from the prior
for each chain.

Fit model:

```{r}
myConfig <- configureMCMC(nycModel)

# block sampler for transmission parameters
paramsForBlock <- c('b', 'w0', 'k')
myConfig$removeSampler(paramsForBlock)
myConfig$addSampler(target = paramsForBlock, type = "RW_block",
                    control = list(adaptInterval = 100,
                                   propCov = diag(c(rep(1, p), 0.5, 0.5))))

myConfig$addMonitors(c('idd_curve', 'reproNum'))

myConfig


myMCMC <- buildMCMC(myConfig)
compiled <- compileNimble(nycModel, myMCMC) 

system.time({
    samples <- runMCMC(compiled$myMCMC, 
                       niter = 100000, nburnin = 80000, 
                       setSeed = c(1, 2, 3),
                       nchains = 3)
})

```

# Check convergence 

Using visual inspection of trace plots

```{r}
paramSamples1 <- samples$chain1[,-grep('idd|reproNum', colnames(samples$chain1))]
paramSamples2 <- samples$chain2[,-grep('idd|reproNum', colnames(samples$chain2))]
paramSamples3 <- samples$chain3[,-grep('idd|reproNum', colnames(samples$chain3))]

par(mfrow = c(2, 5))
for (i in 1:ncol(paramSamples1)) {
    plot(paramSamples1[,i], type = 'l', main = colnames(paramSamples1)[i])
    lines(paramSamples2[,i], col = 'tomato')
    lines(paramSamples3[,i], col = 'dodgerblue')
}
```

# Posterior for IDD curve

```{r}
iddCurveSamples <- rbind(samples$chain1[,grep('idd', colnames(samples$chain1))],
                         samples$chain2[,grep('idd', colnames(samples$chain2))],
                         samples$chain3[,grep('idd', colnames(samples$chain3))])

iddMean <- colMeans(iddCurveSamples)
iddCI <- apply(iddCurveSamples, 2, quantile, probs = c(0.025, 0.975))

iddSummary <- data.frame(time = 1:maxInf, 
                         mean = iddMean,
                         lower = iddCI[1,],
                         upper = iddCI[2,],
                         priorMean = logitDecay(1:maxInf, 4, 1))

ggplot(iddSummary, aes(x = time, y = mean, ymin = lower, ymax = upper)) +
    geom_line(color = 'blue', size = 1) + 
    geom_line(aes(y = priorMean), size = 1) +
    geom_ribbon(fill = 'deepskyblue', alpha = 0.3) + 
    theme_bw() +
    ggtitle('Posterior IDD Transmissibility Curve compared to prior mean')
```

Black line is the prior mean IDD curve.

# Posterior for R0(t)

March 7 - state of emergency declared

March 16 - NYC public schools close

March 17 - NYC bars and restaurants close

March 22 - Stay at home order for non-essential workers

June 8 - Phase 1 reopening

June 22 - Phase 2 reopening

```{r}
r0Samples <- rbind(samples$chain1[,grep('reproNum', colnames(samples$chain1))],
                   samples$chain2[,grep('reproNum', colnames(samples$chain2))],
                   samples$chain3[,grep('reproNum', colnames(samples$chain3))])

r0Mean <- colMeans(r0Samples)
r0CI <- apply(r0Samples, 2, quantile, probs = c(0.025, 0.975))

dates <- nycPeak1$date[-idxStart]

r0Summary <- data.frame(date = nycPeak1$date[-idxStart], 
                        mean = r0Mean,
                        lower = r0CI[1,],
                        upper = r0CI[2,])

ggplot(r0Summary, aes(x = date, y = mean, ymin = lower, ymax = upper)) +
    geom_line(color = 'blue', size = 1) + 
    geom_ribbon(fill = 'deepskyblue', alpha = 0.3) + 
    theme_bw() +
    geom_hline(yintercept = 1, linetype = 2) +
    ylim(0, 9) +
    ggtitle('NYC reproductive number over time')

```


